Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-06 manuscript/manuscript.tex
ms edits and new stuff

A long-standing problem in evolutionary genetics is quantifying the roles of
genetic drift and selection in shaping genome-wide allele frequency changes.
Selection can both directly and indirectly affect allele frequencies, with the
indirect effect coming from the action of selection on correlated loci
elsewhere in genome e.g. linked selection (\cite{Maynard_Smith1974-lc},
\cite{Charlesworth1993-gb,Nordborg1996-nq}; see \cite{Barton2000-zg} for a
review). Previous work on this question has mostly focused on teasing apart
the impacts of drift and selection on genome-wide diversity using population
samples from a single contemporary timepoint, often by modeling the correlation
between regional recombination rate, gene density, and diversity created in the
presence of linked selection \parencite{SellaReview2009,CutterPayseur}. This  %\parencite{Begun1992-ey,Elyashiv2016-vt}
approach has shown linked selection has a major role in shaping patterns of
genome-wide diversity across the genomes of \gc{a range of sexual species}
\parencite{Begun2007-bg,Beissinger2016-cm,Sattath2011-dr,Williamson2014-oy,Andersen2012-bj,Cutter2010-gi,Elyashiv2016-v},
and has allowed us to quantify the relative influence of positive selection
(hitchhiking) and negative selection (background selection;
\cite{Nordborg2005-dc,McVicker2009-ax,Hernandez2011-gs,Elyashiv2016-vt}.
However, we lack an understanding of how genome-wide linked selection acts over
time.

There are numerous examples of rapid phenotypic adaptation
\parencite{Grant2011-wk,Grant2006-hj,Reznick1997-mh,Franks2007-dr} and rapid
\gc{, selection-driven} genomic evolution in asexual populations
\parencite{Good2017-om,Bennett1990-bc,Baym2016-kh}.  Yet the polygenic nature
of fitness makes detecting the impact selection on genome-wide variation over
short timescales in sexual populations remarkably difficult. This is because
the effect of selection on a polygenic trait (such as fitness) is distributed
across loci in proportion to their effect sizes. This can lead to subtle allele
frequency shifts on standing variation that are difficult to distinguish from
background levels of genetic drift and sampling variance.  However,
increasingly genomic experimental evolution studies with multiple timepoints,
and in some cases multiple replicate populations, are being used to detect
\gc{large effect} selected loci \parencite{Turner2011-sx,Turner2012-bm} and differentiate modes
of selection \parencite{Burke2010-tz,Barghi2019-qy,Therkildsen2019-zy}.  In
addition these temporal-genomic studies have begun in wild populations, some
with the goal of finding variants that exhibit frequency changes consistent
with fluctuating selection \parencite{Bergland2014-ij,Machado2018-cs}. In a
previous paper, we proposed that one useful signal for understanding the
genome-wide impact of polygenic linked selection detectable from temporal
genomic studies is the temporal autocovariance in allele frequency changes
\parencite{Buffalo2019-io}.  These covariances are directly estimable from
temporal genomic data and are created when the loci that underly heritable
fitness variation perturb the frequencies of linked neutral alleles; in
contrast, when genetic drift acts alone in a closed population, these
covariances are zero in expectation. Mathematically, temporal covariances are
useful because it is natural to decompose the total variance in allele
frequency change across a set of time intervals into the variances and
covariances in allele frequency change among time intervals. Furthermore,
biologically, these covariances reflect the extent to which allele frequency
changes in one generation predict changes in another due to a shared \gc{genetic} %fitness
background and \gc{correlated} selection pressures. %shared

Here, we provide the first empirical analyses to quantify the impact of linked
selection acting over short timescales (tens of generations) across two evolve
and re-sequence studies \parencite{Barghi2019-qy,Kelly2019-dc}, and \gc{an}
artificial selection experiment \parencite{Castro2019-uk}. We repeatedly find a
signal of temporal covariance, consistent with linked selection acting to
significantly perturb genome-wide allele frequency changes across the genome in
a manner that other approaches would not be able differentiate from genetic
drift. We estimate the lower bound on the proportion of total variation in
allele frequency change caused by \gc{selection}, and the correlation
between  %linked
allele frequency changes between replicate populations caused by response to
convergent selection pressures. Overall, we demonstrate that linked selection
has a powerful role in shaping genome-wide allele frequency changes over very
short timescales.

@@

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to estimate the quantify
the impact of linked selection in have \gc{found over obligately
  sexual origanisms SHOULD WE SAY 0-89} 19\% to 89\% \vb{reduction
in genome-wide diversity} over long time periods is driven by \vb{linked
selection}
\parencite{McVicker2009-ax,Elyashiv2016-vt,Corbett-Detig2015-gt,Coop2016-gx,Comeron2014-nh}
Thus there is reason to be hopeful that we could detect this from temporal
data, which would help to resolve the timescales that linked selection act
over. In our reanalysis of the \textcite{Barghi2019-qy} study, we find evidence
of complex linked selection dynamics, with selection pressures flipping over
time due to either environmental change, the breakup of epistatic combinations
or advantageous haplotypes. Such patterns would be completely obscured in
samples only from contemporary populations, thus we can hope to have a much
richer picture of linked selection as temporal sequencing becomes more common,
\vb{perhaps allowing us to observe the effects of ecological dynamics in
genomic data \parencite{Hairston2005-ga}.}

@@

\subsection{Across-Replicate Covariances}

Here, we extend our theoretic work in \textcite{Buffalo2019-io} to
between-replicate allele frequency change covariances. We focus on the
between-replicate covariance at a focal neutral site across two replicate
populations, $A$ and $B$. As in \textcite{Buffalo2019-io}, the heritable allele
frequency change (which is the only term that contributes to the covariance),
can be written as a linear combination of effect sizes $\alpha_{t,l,A}$ and LD
between the neutral site and selected site $l$, $D_{t,l,A}$.

\begin{align}
  \cov(\Delta p_{t,A}, \Delta p_{t,B}) = \cov\left(\sum_l \alpha_{t,A,l} D_{t,A,l}, \sum_l \alpha_{t,B,l} D_{t,B,l}\right)
\end{align}

Since the replicate populations derive from a common population at time $t=0$
before their split, they share a fraction of their LD plus a replicate-specific
deviation due to replicate-specific selection and drift. Thus, we can partition
the $D$ terms as: $D_{t,A,l} = D_{0,l} + \varepsilon_{t,A}$ (and likewise for
replicate $B$). Under drift alone, this deviation can be found from equation
(4) of \textcite{Ragsdale2019-fs}.  Similarly, in the original population the
additive effects at each locus were identical ($\alpha_{0,l}$), but in each
replicate population have since been perturbed in a replicate-specific manner
(the term $a_{t,A,l}$) as well as a change induced by convergent selection
pressure after $t$ generations, $b_{t,l}$, $\alpha_{0,l} + b_{t,l} + a_{t,A,l}$
(and likewise for replicate $B$).

We then can write,

\begin{align}
  \cov(\Delta p_{t,A}, \Delta p_{t,B}) = \cov\left(\sum_l \alpha_{t,A,l} (D_{0,l} + \varepsilon_{A,l}), \sum_l \alpha_{t,B,l} (D_{0,l} + \varepsilon_{B,l})\right)
\end{align}

We define a statistic similar to $G$ for estimating the proportion of allele
frequency change common between two replicate populations due to linked
selection. Covariance in allele frequency change between two replicate
populations is due to convergent selection pressure selecting haplotypes shared
between the two replicate populations, which acts to perturb linked neutral
variation in parallel way. Similar to \textcite{Buffalo2019-io} equation (1).
We can similarly partition the allele frequency change,

\begin{align}
  \label{eq:shared-partition}
  \Delta p_t &= \underbrace{\Delta_{_N} p_t  + \Delta_{_M} p_t}_\text{drift} + \underbrace{\Delta_{_S} p_t}_\text{shared} + \underbrace{\Delta_{_U} p_t}_\text{unique}
\end{align}

where $\Delta_{_N} p_t$ and $\Delta_{_M} p_t$ are the allele frequency changes
due to non-heritable variation in fitness (e.g. environmental differences in
family size) and Mendelian segregation. In \textcite{Buffalo2019-io} allele
frequency changes due to heritable fitness variation are captured in the term
$\Delta_{_H} p_t$; here we further partition this into heritable frequency
changes shared between replicates ($\Delta_{_S} p_t$) and those unique to a
replicate ($\Delta_{_H} p_t$). This is important, as we can

We base this partition on our model in
\textcite{Buffalo2019-io}, where we decompose the LD between the 
focal neutral site and selected sites conditioning on whether

\begin{align}
  \Delta_{_H} p_{t,A} &= \sum_l \alpha_{t,l,A} D_{t,l, A} \\
                      &= \alpha_{t,l,A} D_{t,l,A} + (\alpha_{t,l,A'} + \alpha_{t,l,S}) D_{t,l,S} \\
\end{align}

By definition,

\begin{align}
  \cov(\Delta p_{s,A}, \Delta p_{t,B}) &= \cov(\Delta_{_S} p_{s,A}, \Delta_{_S} p_{t,B}), \;\; A \ne B\\
  \cov(\Delta p_{s,A}, \Delta p_{t,A}) &= \cov(\Delta_{_S} p_{s,A} + \Delta_{_U} p_{s,A}, \Delta_{_S} p_{t,A} + \Delta_{_U} p_{t,A}) 
\end{align}

Then, we define a replicate covariance analog of $G$,

@@

For selection studies with a control line (e.g. the \cite{Castro2019-uk} study
we have reanalyzed), we have an alternate approach to quantify the effect of
linked selection: we can compare the total variance in allele frequency change
in control and selection lines. By the orthogonality of the components in
equation XXX,

\begin{align}
  \var(\Delta p_t - p_0) = \sum_t \var(\Delta {_N} p_t ) + \sum_t \var(\Delta {_M} p_t ) + 
  \var\left(\sum_t \Delta {_S} p_t\right) + \var\left(\sum_t \Delta {_U} p_t\right)
\end{align}

the term $\sum_t \var(\Delta {_S} p_t)$

\begin{align}
  \var\left(\sum_t \Delta_{_S} p_t\right) = 
\sum_t \var( \Delta_{_U} p_t) + \sum_{i \ne j} \cov( \Delta_{_S} p_i, \Delta_{_S} p_j)
\end{align}

In the \textcite{Castro2019-uk} study, we 
quantify the effect of linked selection as:

\begin{align}
  \frac{1}{2} \left( \var(\Delta p_{20, \mathrm{LS1}} - p_{0, \mathrm{LS1}}) + \var(\Delta p_{20, \mathrm{LS2}} - p_{0, \mathrm{LS2}}) \right) -  \var(\Delta p_{20, \mathrm{Ctrl}} - p_{0, \mathrm{Ctrl}}) = \\
  \cov(p_{20, \mathrm{LS1}} - p_{0, \mathrm{LS1}}, p_{20, \mathrm{LS2}} - p_{0, \mathrm{LS2}})
\end{align}

>>>>

A long-standing problem in evolutionary genetics is quantifying the roles of
genetic drift and selection in shaping genome-wide allele frequency changes.
Selection can both directly and indirectly affect allele frequencies, with the
indirect effect coming from the action of selection on correlated loci
elsewhere in genome e.g. linked selection (\cite{Maynard_Smith1974-lc},
\cite{Charlesworth1993-gb,Nordborg1996-nq}; see \cite{Barton2000-zg} for a
review). Previous work on this question has mostly focused on teasing apart
the impacts of drift and selection on genome-wide diversity using population
samples from a single contemporary timepoint, often by modeling the correlation
between regional recombination rate, gene density, and diversity created in the
presence of linked selection \parencite{Begun1992-ey,Elyashiv2016-vt}. This
approach has shown linked selection has a major role in shaping patterns of
genome-wide diversity across the genomes of some of organisms
\parencite{Begun2007-bg,Beissinger2016-cm,Sattath2011-dr,Williamson2014-oy,Andersen2012-bj,Cutter2010-gi},
and has allowed us to quantify the relative influence of positive selection
(hitchhiking) and negative selection (background selection;
\cite{Nordborg2005-dc,McVicker2009-ax,Hernandez2011-gs,Elyashiv2016-vt}.
However, we lack an understanding of how genome-wide linked selection acts over
time.

There are numerous examples of rapid phenotypic adaptation
\parencite{Grant2011-wk,Grant2006-hj,Reznick1997-mh,Franks2007-dr} and rapid
genomic evolution in asexual populations
\parencite{Good2017-om,Bennett1990-bc,Baym2016-kh}.  Yet the polygenic nature
of fitness makes detecting the impact selection on genome-wide variation over
short timescales in sexual populations remarkably difficult. This is because
the effect of selection on a polygenic trait (such as fitness) is distributed
across loci in proportion to their effect sizes. This can lead to subtle allele
frequency shifts on standing variation that are difficult to distinguish from
background levels of genetic drift and sampling variance.  However,
increasingly genomic experimental evolution studies with multiple timepoints,
and in some cases multiple replicate populations, are being used to detect
selected loci \parencite{Turner2011-sx,Turner2012-bm} and differentiate modes
of selection \parencite{Burke2010-tz,Barghi2019-qy,Therkildsen2019-zy}.  In
addition these temporal-genomic studies have begun in wild populations, some
with the goal of finding variants that exhibit frequency changes consistent
with fluctuating selection \parencite{Bergland2014-ij,Machado2018-cs}. In a
previous paper, we proposed that one useful signal for understanding the
genome-wide impact of polygenic linked selection detectable from temporal
genomic studies is the temporal autocovariance in allele frequency changes
\parencite{Buffalo2019-io}.  These covariances are directly estimable from
temporal genomic data and are created when the loci that underly heritable
fitness variation perturb the frequencies of linked neutral alleles; in
contrast, when genetic drift acts alone in a closed population, these
covariances are zero in expectation. Mathematically, temporal covariances are
useful because it is natural to decompose the total variance in allele
frequency change across a set of time intervals into the variances and
covariances in allele frequency change among time intervals. Furthermore,
biologically, these covariances reflect the extent to which allele frequency
changes in one generation predict changes in another due to a shared fitness
background and shared selection pressures.

Here, we provide the first empirical analyses to quantify the impact of linked
selection acting over short timescales (tens of generations) across two evolve
and re-sequence studies \parencite{Barghi2019-qy,Kelly2019-dc}, and one
artificial selection experiment \parencite{Castro2019-uk}. We repeatedly find a
signal of temporal covariance, consistent with linked selection acting to
significantly perturb genome-wide allele frequency changes across the genome in
a manner that other approaches would not be able differentiate from genetic
drift. We estimate the lower bound on the proportion of total variation in
allele frequency change caused by linked selection, and the correlation between
allele frequency changes between replicate populations caused by response to
convergent selection pressures. Overall, we demonstrate that linked selection
has a powerful role in shaping genome-wide allele frequency changes over very
short timescales.

@@

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to estimate the quantify
the impact of linked selection in have suggested a 19\% to 89\% \vb{reduction
in genome-wide diversity} over long time periods is driven by \vb{linked
selection}
\parencite{McVicker2009-ax,Elyashiv2016-vt,Corbett-Detig2015-gt,Coop2016-gx,Comeron2014-nh}
Thus there is reason to be hopeful that we could detect this from temporal
data, which would help to resolve the timescales that linked selection act
over. In our reanalysis of the \textcite{Barghi2019-qy} study, we find evidence
of complex linked selection dynamics, with selection pressures flipping over
time due to either environmental change, the breakup of epistatic combinations
or advantageous haplotypes. Such patterns would be completely obscured in
samples only from contemporary populations, thus we can hope to have a much
richer picture of linked selection as temporal sequencing becomes more common,
\vb{perhaps allowing us to observe the effects of ecological dynamics in
genomic data \parencite{Hairston2005-ga}.}

@@

\subsection{Replicate $G$ and Partitioning the Variance in Allele Frequency}
\label{supp:replicate-g}

We define a statistic similar to $G$ for estimating the proportion of allele
frequency change common between two replicate populations due to linked
selection. Covariance in allele frequency change between two replicate
populations is due to convergent selection pressure selecting haplotypes shared
between the two replicate populations, which acts to perturb linked neutral
variation in parallel way.

@@

Extending our theoretic work in \textcite{Buffalo2019-io}, we can partition the
allele frequency change in two replicates into drift, and shared selection and
replicate-specific selection components of allele frequency change. For two
replicates, $A$ and $B$,

\begin{align}
  \Delta p_{t,A} &= \Delta_{_D} p_{t,A} + \Delta_{_{U}} p_{t,A} + \Delta_{_S} p_t \\
  \Delta p_{t,B} &= \Delta_{_D} p_{t,B} + \Delta_{_{U}} p_{t,B} + \Delta_{_S} p_t
\end{align}

where $\Delta_{_D} p_{t,A}$ is allele frequency change due to drift (this is
specific to a replicate, and equal to $\Delta_{_N} p_{t,A} + \Delta_{_M}
p_{t,A}$ in the notation of \cite{Buffalo2019-io}), $\Delta_{_{U}} p_{t,A}$ is
the allele frequency change from indirect selection specific to replicate $A$
(and likewise with $\Delta_{_{U}} p_{t,A}$ for replicate $B$), and $\Delta_{_S}
p_t$ is the allele frequency change from indirect selection shared across the
replicates $A$ and $B$ (this term lacks a replicate subscript since by
construction it is identical between replicates). By construction, each of
these terms is uncorrelated, so the variance and be written as:

\begin{align}
  \var(\Delta p_{t,A}) &= \var(\Delta_{_D} p_{t,A}) + \var(\Delta_{_{U}} p_{t,A}) + \var(\Delta_{_S} p_t) \\
\end{align}

The shared effects of indirect selection can be quantified from the observed
allele frequency changes, since the covariance in allele frequency change
across replicates is the covariance of the shared term by construction,

\begin{align}
  \cov(\Delta p_{t,A}, \Delta p_{t,B}) = \cov(\Delta_{_S} p_{t}, \Delta_{_S} p_{t}) = \var(\Delta_{_S} p_{t})
\end{align}

In artificial selection studies with a control line, such as the
\textcite{Castro2019-uk} study, this allows us to estimate the contribution of
the effects of shared and unique indirect selection.

\begin{align}
  \var(\Delta p_{0,\mathrm{Ctrl}}) = \var()
\end{align}

% Here, we extend our theoretic work in \textcite{Buffalo2019-io} to
% between-replicate allele frequency change covariances. We focus on the
% between-replicate covariance at a focal neutral site across two replicate
% populations, $A$ and $B$. As in \textcite{Buffalo2019-io}, the heritable allele
% frequency change (which is the only term that contributes to the covariance),
% can be written as a linear combination of effect sizes $\alpha_{t,l,A}$ and LD
% between the neutral site and selected site $l$, $D_{t,l,A}$.

% \begin{align}
%   \cov(\Delta p_{t,A}, \Delta p_{t,B}) = \cov\left(\sum_l \alpha_{t,A,l} D_{t,A,l}, \sum_l \alpha_{t,B,l} D_{t,B,l}\right)
% \end{align}

% Since the replicate populations derive from a common population at time $t=0$
% before their split, they share a fraction of their LD plus a replicate-specific
% deviation due to replicate-specific selection and drift. Thus, we can partition
% the $D$ terms as: $D_{t,A,l} = D_{0,l} + \varepsilon_{t,A}$ (and likewise for
% replicate $B$). Under drift alone, this deviation can be found from equation
% (4) of \textcite{Ragsdale2019-fs}.  Similarly, in the original population the
% additive effects at each locus were identical ($\alpha_{0,l}$), but in each
% replicate population have since been perturbed in a replicate-specific manner
% (the term $a_{t,A,l}$) as well as a change induced by convergent selection
% pressure after $t$ generations, $b_{t,l}$, $\alpha_{0,l} + b_{t,l} + a_{t,A,l}$
% (and likewise for replicate $B$).

% We then can write,

% \begin{align}
%   \cov(\Delta p_{t,A}, \Delta p_{t,B}) = \cov\left(\sum_l \alpha_{t,A,l} (D_{0,l} + \varepsilon_{A,l}), \sum_l \alpha_{t,B,l} (D_{0,l} + \varepsilon_{B,l})\right)
% \end{align}

% We define a statistic similar to $G$ for estimating the proportion of allele
% frequency change common between two replicate populations due to linked
% selection. Covariance in allele frequency change between two replicate
% populations is due to convergent selection pressure selecting haplotypes shared
% between the two replicate populations, which acts to perturb linked neutral
% variation in parallel way. Similar to \textcite{Buffalo2019-io} equation (1).
% We can similarly partition the allele frequency change,

% \begin{align}
%   \label{eq:shared-partition}
%   \Delta p_t &= \underbrace{\Delta_{_N} p_t  + \Delta_{_M} p_t}_\text{drift} + \underbrace{\Delta_{_S} p_t}_\text{shared} + \underbrace{\Delta_{_U} p_t}_\text{unique}
% \end{align}

% where $\Delta_{_N} p_t$ and $\Delta_{_M} p_t$ are the allele frequency changes
% due to non-heritable variation in fitness (e.g. environmental differences in
% family size) and Mendelian segregation. In \textcite{Buffalo2019-io} allele
% frequency changes due to heritable fitness variation are captured in the term
% $\Delta_{_H} p_t$; here we further partition this into heritable frequency
% changes shared between replicates ($\Delta_{_S} p_t$) and those unique to a
% replicate ($\Delta_{_H} p_t$). This is important, as we can

% We base this partition on our model in
% \textcite{Buffalo2019-io}, where we decompose the LD between the 
% focal neutral site and selected sites conditioning on whether

% \begin{align}
%   \Delta_{_H} p_{t,A} &= \sum_l \alpha_{t,l,A} D_{t,l, A} \\
%                       &= \alpha_{t,l,A} D_{t,l,A} + (\alpha_{t,l,A'} + \alpha_{t,l,S}) D_{t,l,S} \\
% \end{align}

% By definition,

% \begin{align}
%   \cov(\Delta p_{s,A}, \Delta p_{t,B}) &= \cov(\Delta_{_S} p_{s,A}, \Delta_{_S} p_{t,B}), \;\; A \ne B\\
%   \cov(\Delta p_{s,A}, \Delta p_{t,A}) &= \cov(\Delta_{_S} p_{s,A} + \Delta_{_U} p_{s,A}, \Delta_{_S} p_{t,A} + \Delta_{_U} p_{t,A}) 
% \end{align}

% For selection studies with a control line (e.g. the \cite{Castro2019-uk} study
% we have reanalyzed), we have an alternate approach to quantify the effect of
% linked selection: we can compare the total variance in allele frequency change
% in control and selection lines. By the orthogonality of the components in
% equation XXX,

% \begin{align}
%   \var(\Delta p_t - p_0) = \sum_t \var(\Delta {_N} p_t ) + \sum_t \var(\Delta {_M} p_t ) + 
%   \var\left(\sum_t \Delta {_S} p_t\right) + \var\left(\sum_t \Delta {_U} p_t\right)
% \end{align}

% the term $\sum_t \var(\Delta {_S} p_t)$

% \begin{align}
%   \var\left(\sum_t \Delta_{_S} p_t\right) = 
% \sum_t \var( \Delta_{_U} p_t) + \sum_{i \ne j} \cov( \Delta_{_S} p_i, \Delta_{_S} p_j)
% \end{align}

% In the \textcite{Castro2019-uk} study, we 
% quantify the effect of linked selection as:

% \begin{align}
%   \frac{1}{2} \left( \var(\Delta p_{20, \mathrm{LS1}} - p_{0, \mathrm{LS1}}) + \var(\Delta p_{20, \mathrm{LS2}} - p_{0, \mathrm{LS2}}) \right) -  \var(\Delta p_{20, \mathrm{Ctrl}} - p_{0, \mathrm{Ctrl}}) = \\
%   \cov(p_{20, \mathrm{LS1}} - p_{0, \mathrm{LS1}}, p_{20, \mathrm{LS2}} - p_{0, \mathrm{LS2}})
% \end{align}
